---
title: "BUSN41204 Week 1 Review Session"
author: "Aman Krishna"
date: "1/7/2022"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Welcome to the first review session for BUSN41204! Today we will go over the 
Philly Crime dataset, which appeared in a previous midterm. We will stick with R
for demonstration.

*The write-up is available on Canvas in both `.pdf` and `.Rmd`. In order to replicate the results yourself, please also download the `PhillyCrime.csv` dataset from Canvas.*

## Free Resources for Learning R

Here are some free online resources for learning R:

- [Codecademy](https://www.codecademy.com/learn/learn-r)

- [R for Beginners](https://cran.r-project.org/doc/contrib/Paradis-rdebuts_en.pdf)
by Emmanuel Paradis.

- [Swirl](https://swirlstats.com), an interactive program for learning R.

- [Additional Resources](https://chicagoboothml.github.io/MLClass/computing/).

## Ed

- Please post any question you may have on the [Ed](https://edstem.org)
  platform. You will probably get a faster response from the TAs, Professor Kolar,
  or one of your peers. Making your questions public can also benefit your
  classmates who have similar questions.
  
# Getting Started

- What is R?

  - R is an open-source, free programming language popular in data science.
  
  - R has many packages, code written by other users that implement classic
  machine learning algorithms, such as kNN, Lasso.
  
  - R is user-friendly and easy to pickup.

- Why R instead of Python?

  - Python users may use the online [Jupyter Notebook Server](https://jupyter-class.chicagobooth.edu/) 
    instead.
    
  - Python: swiss army knife. R: letter opener. Sometimes you don't need every
  bell and whistle.

- What is R Studio?

  - An interface for viewing R code, variables, and plots. All in the same place.
  
  - We *strongly recommend* using the [R Studio Server](https://rstudio-class.chicagobooth.edu)
    instead of installing locally. Please login using your *CNetID*, rather than
    BoothID.
    
  - But if you want, you can also download RStudio [here](http://www.rstudio.com/products/rstudio/download)
    and download R [here](http://cran.r-project.org).
    
    - Setting up R on your local device can be time consuming (Windows users may
    have to deal with environmental variables). 

- What is R Markdown?

  - A package `rmarkdown` + Rstudio + latex: converts your code and writeup into
    pdf, or html, or presentation.
    
  - You may use R Markdown to prepare your homework submissions. (For learning 
    markdown, [here](https://www.markdownguide.org/getting-started/) is a great 
    resource. R Markdown is basically markdown with chunks of R code.) A word of
    caution: R Markdown takes time to get used to and please use R Markdown 
    *only if* you feel like spending the extra time to learn it.
    
  - Please feel free to use Word and paste the R outputs to the word
    doc. In most cases, preparing a document where you paste results from R will be
    sufficient.
    
  - R Markdown tutorials can be found 
    [here](http://chicagoboothml.github.io/MachineLearning_Fall2015/Tutorials/R%20Markdown%20Tutorial/)
    and [here](https://bookdown.org/yihui/rmarkdown/).
  
  - A template can be found 
    [here](https://raw.githubusercontent.com/ChicagoBoothML/ML2016/master/code/BostonHousing_KNN_BiasVarTradeOff_CrossValid.Rmd).
  
  - Again, you *do not* have to use R Markdown.
  
  - Setting up R Markdown on your local device can also be time consuming.

## Installing and using packages in R
Use the function `install.packages()` and type the name of the package you want
to install, *with quotation marks*, between the parenthesis.
```{r, eval=FALSE}
install.packages('swirl')
```

**DO NOT** include the `install.packages()` function in your
R markdown writeup. Otherwise your Rmarkdown file won't compile
(cannot be knit into pdf).

For using packages, use the `library()` function.
```{r}
library('swirl')
```

A more detailed guide for installing packages in R can be found [here](http://www.sthda.com/english/wiki/installing-and-using-r-packages).

If you wish to install R locally, you may need to run the following script to
install the required packages for your local machine. *The R Studio on the server
comes with all these packages installed.*
```{r, eval=FALSE}
packageNames = c("MASS", "ISLR", "animation", "ElemStatLearn", "glmnet", 
                 "textir", "nnet", "methods", "statmod", "stats", "graphics", 
                 "RCurl", "jsonlite", "tools", "utils", "data.table", "gbm", 
                 "ggplot2", "randomForest", "tree", "class", "kknn", "e1071",
                 "data.table", "recommenderlab")

for (pkgName in packageNames) {
  if (!(pkgName %in% rownames(installed.packages()))) {
    install.packages(pkgName,
                     dependencies=TRUE,
                     repos='http://cran.rstudio.com')
  }
}

update.packages(ask=FALSE)
```

## Setting Working Directory

A good practice is to keep all the files associated with a project in separate
directories (folders). It is usually a good idea to make a separate folder for
each homework assignment, final project, review session, etc. You can browse the
files you have on the `Files` tab, bottom-right window.

You can use the following command to tell R that you are working in a specific
directory.
```{r}
setwd('~/w23 review session 1')
```

Alternatively, click the folder for the project you want and then click `More`
(a small blue gear icon). Select `Set As Working Directory` from the drop down
menu.

You can verify that your working directory is setup correctly by:
1. Running the following command
```{r}
getwd()
```
2. Go the the `Files` tab in the bottom-right window. Click `More` and select
`Go To Working Directory` from the drop down menu.

## Uploading Files to R Studio Server

1. Open up the R Studio webpage.

2. Go to the bottom right window and click on the `Files` tab.

3. Click `Upload`. It looks like a whiteboard with a tiny bronze "up" button on
the top left corner.

Let's do that for the `PhillyCrime` dataset which we will later use!

## Getting Help

- Again, please use [Ed](https://edstem.org)!
- You can also view the documentation using the command `help()`
```{r}
help(swirl)
help(length)
```
- Alternatively, Google it! There are a lot of amazing free resources online 
  that could probably answer your questions better than the TAs can.

\newpage

# Crimes in Philly
The dataset `PhillyCrime.csv` contains information of two types of crime
incidents (`Vandalism` and `Thefts`) occurred in Philadelphia in 2016^[The whole
dataset is available at https://www.opendataphilly.org/dataset/crime-incidents]. 

In this problem, we will use $k$ nearest neighbor to predict the category of the
crimes^[Code for this problem is shamelessly borrowed from Jianeng Xu's writeup
from a previous iteration of the course].

1. Make two separate scatterplots of the crime incidents using their latitudes
(`X`) and longitudes (`Y`), one for `Vandalism` and one for `Thefts`. 
Describe any pattern or difference you find from the scatterplots.
(You might want to decrease the size of the points so that they don’t overlap
too much.)

Downloading helper files and importing required libraries.
```{r}
set.seed(100)
download.file("https://raw.githubusercontent.com/ChicagoBoothML/HelpR/master/docv_classification.R",
              destfile = "docv_classification.R")
source("docv_classification.R")
PhillyCrime <- read.csv("PhillyCrime.csv")
library(MASS) 
library(kknn)
```


```{r}
par(mfrow = c(1,2))
thefts = PhillyCrime[PhillyCrime$Category=="Thefts",]
plot(thefts$X, thefts$Y, cex = 0.1, xlab = "X", ylab = "Y", 
     main = "Thefts", col = "blue")
vandalism = PhillyCrime[PhillyCrime$Category=="Vandalism",]
plot(vandalism$X, vandalism$Y, cex = 0.1, xlab = "X", ylab = "Y", 
     main = "Vandalisim", col = "blue")
```
<!-- If you have trouble running the code above on the R studio server, please check -->
<!-- the Ed thread [here](https://edstem.org/us/courses/15788/discussion/998679). -->

2. Split the data, with 50% into a training set and the other 50% reserved for a
   validation set. Fit a set of $k$-nearest neighbors classifiers for 
   $k = 1, 2, . . . , 100$ on the training set to predict the crime category 
   using latitude (`X`) and longitude (`Y`) as predictors.
   
    1. Make a scatterplot of the misclassification rate on the validation set 
       against the parameter $k$.
       
```{r message=FALSE}
compute_error = function(train, test, k_vec){
  error_vec = c()
  for(i in 1:length(k_vec)){
    fit = kknn(Category ~ X + Y, train, test, k = k_vec[i],
               kernel = "rectangular")
    results = table(fit$fitted.values, test$Category)
    print(results)
    error_vec[i] = (results[1,2] + results[2,1]) / dim(test)[1]
  }
  return(error_vec)
}
```

    If you have problem running `kknn` and R spits out:
    `Error in if (response != "continuous") { : argument is of length zero`,
    please run the following code chunk
```{r}
PhillyCrime$Category = as.factor(PhillyCrime$Category)
```
    It tells R to treat the variable `Category` as a categorical variable, rather
    than a string. 

```{r message=FALSE, results='hide',fig.keep='all'}
# Splitting the data into 50% training 50% testing.
N = dim(PhillyCrime)[1]  # Number of samples
train_ratio = 0.5
train_index = sample(N,train_ratio*N)
train = PhillyCrime[train_index,]
test = PhillyCrime[-train_index,]
# We only focus on k = 1 to 40 to save time
k_vec =  1:40
error_vec = compute_error(train, test, k_vec)
plot(k_vec, error_vec)
```
    
    2. Report the optimal $k$ selected by the validation-set approach and the
       minimum misclassification rate on the validation set.
       
```{r}
best_k = k_vec[which.min(error_vec)]
best_k
```
      
    3. Make a single scatterplot of the crime incidents using their latitudes 
       (`X`) and longitudes (`Y`). Color the points according to their predicted
       class given by the optimal $k$-nearest neighbors selected above.

```{r}
knn = kknn(Category~X+Y,train,test, k=best_k, kernel = "rectangular") 
plot(test$X, test$Y, col = as.numeric(knn$fitted.values), cex = 0.1)
legend("bottomright", c("Thefts","Vandalism"), col = c(1,2), pch = c(1,1))
```
    
3. Repeat (2) for 20 times (with 20 random splits of the dataset).
```{r message=FALSE, results='hide',fig.keep='all'}
train_ratio = 0.5
error_mat = c()
for (j in 1:2) {
  train_index = sample(N,train_ratio*N)
  train = PhillyCrime[train_index,]
  test = PhillyCrime[-train_index,]
  k_vec = 1:40
  error_vec_temp = compute_error(train, test, k_vec)
  error_mat = rbind(error_mat, error_vec_temp)
}

best_k_1 = c()
min_error_1 = c()

for (i in 1:dim(error_mat)[1]) {
  best_k_1[i] = k_vec[which.min(error_mat[i,])]
  min_error_1[i] = min(error_mat[i,])
}
```

    1. Report the 20 optimal $k$’s selected by the validation sets.

```{r}
best_k_1
```

    2. Report the average of the minimum out-of-sample misclassification rates 
       as well as its standard error.
       
       The minimal misclassification rate for each split is
       
```{r}
min_error_1
```

       The average and standard error of minimal misclassification rate are

```{r}
mean(min_error_1)
```

       and 
       
```{r}
sd(min_error_1)
```
    
4. Repeat (2) and (3), this time with a 90/10 split of the data. Make sure you
include in your solutions

    1. (For a particular split,) a scatterplot of the misclassification rate on
     the validation set against the parameter $k$.
      
```{r message=FALSE,results='hide',fig.keep='all'}
train_ratio = 0.9
N = dim(PhillyCrime)[1]
train_index = sample(N,train_ratio*N)
train = PhillyCrime[train_index,]
test = PhillyCrime[-train_index,]
k_vec = 1:40
error_vec2 = compute_error(train, test, k_vec)
plot(k_vec, error_vec2)
```
      
    2. (For a particular split,) a scatterplot of the crime incidents (`Y` vs
       `X`) with points colored according to predicted class given by the 
       optimal $k$.
      
```{r}
best_k = k_vec[which.min(error_vec2)]
best_k
```
The scatterplot can be seen below.
```{r}
knn = kknn(Category~X+Y,train,test, k=best_k, kernel = "rectangular") 
plot(test$X, test$Y, col = as.numeric(knn$fitted.values), cex = 0.1) 
legend("bottomright", c("Thefts","Vandalism"), col = c(1,2), pch = c(1,1))
```


    3. 20 optimal $k$’s selected by the validation sets.

```{r, message=FALSE, results='hide'}
train_ratio = 0.9 
error_mat_2 = c() 
for(j in 1:20){ 
  train_index = sample(N,train_ratio * N) 
  train = PhillyCrime[train_index,] 
  test = PhillyCrime[-train_index,]
  k_vec = 1:40 
  error_vec_temp = compute_error(train, test, k_vec) 
  error_mat_2 = rbind(error_mat_2, error_vec_temp) 
  }

best_k_2 = c() 
min_error_2 = c() 
for(i in 1:dim(error_mat_2)[1]){
  best_k_2[i] = k_vec[which.min(error_mat_2[i,])]
  min_error_2[i] = min(error_mat_2[i,]) 
  }
```

    4. Average minimum out-of-sample misclassification rates and its standard
       error.
  
```{r}
print(best_k_2)

print(min_error_2)
# average minimum out-of-sample misclassification rates
print(mean(min_error_2))
# standard error of out-of-sample misclassification rates
print(sd(min_error_2))

```
    
5. Comment on the difference between the results obtained in (2), (3) and (4).

    - We repeat validation multiple times in question 2.3. We can see that the 
      optimal k varies across different random splits, which is due to us 
      randomly splitting the testing sets.
    
    - In question 2.4 (10 percent testing set), the average of optimal k is 
      slightly higher than that of 50/50 case (question 2.3). Because in Q2.4 
      the training set is bigger, there are more data points in the neighbors.
      
    - Also in question 2.4 we get smaller average error but larger variance of 
      errors across different runs. The lower mean is due to larger training set
      because there are more data in the neighbors of testing set, which might
      increase prediction accuracy. The bigger variance is due to smaller size 
      of testing set.

6. For a 25-nearest neighbors classifier trained on 50% of the data in 
   `PhillyCrime.csv`, plot the ROC curve calculated on a validation set with 
   the other 50% of the data.
   
```{r message=FALSE}
library(ROCR)
train_ratio = 0.5 
train_index = sample(N,train_ratio * N) 
train = PhillyCrime[train_index,] 
test = PhillyCrime[-train_index,]
knn = kknn(Category~X+Y,train,test, k=25, kernel = "rectangular") 
pred = prediction(as.numeric(knn$fitted.values), as.numeric(test$Category)) 
perf = performance(pred, measure = "tpr", x.measure = "fpr") 
plot(c(0,1),c(0,1),xlab='False Positive Rate', ylab='True Positive Rate',main="ROC curve",cex.lab=1,type="n") 
lines(perf@x.values[[1]], perf@y.values[[1]])
```
