---
title: "Tree Classification Example"
author: ""
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

A young bank wants to explore ways of converting its 
liability (deposit) customers to personal loan customers.
A campaign the bank ran for liability customers showed a healthy conversion
rate of over 9% successes. This has encouraged the retail marketing department to
devise smarter campaigns with better target marketing. The goal of our analysis
is to model the previous campaign’s customer behavior to analyze what combination 
of factors make a customer more likely to accept a personal loan. This
will serve as the basis for the design of a new campaign.

Column Name                 Description
------------------          ----------------
ID	                        Customer ID
Age	                        Customer's age in completed years
Experience	                #years of professional experience
Income	                    Annual income of the customer ($000)
ZIPCode	                    Home Address ZIP code.
Family	                    Family size of the customer
CCAvg	                      Avg. spending on credit cards per month ($000)
Education	                  Education Level. 1: Undergrad; 2: Graduate; 3: Advanced/Professional
Mortgage                    Value of house mortgage if any. ($000)
Personal Loan	              Did this customer accept the personal loan offered in the last campaign?
Securities Account	        Does the customer have a securities account with the bank?
CD Account	                Does the customer have a certificate of deposit (CD) account with the bank?
Online	                    Does the customer use internet banking facilities?
CreditCard	                Does the customer use a credit card issued by UniversalBank?



### Analysis

Loading libraries 
```{r}
library(rpart)
library(rpart.plot)
library(caret)    # used for confusionMatrix
```

Loading data
```{r}
bank.df <- read.csv("UniversalBank.csv")
dim(bank.df)
```


Drop ID and zip code columns.
```{r}
bank.df <- bank.df[ , -c(1, 5)] 
bank.df$Personal.Loan <- as.factor(bank.df$Personal.Loan)
```

Partition 5000 observations randomly into training (3000 records) and validation (2000 records).
```{r}
set.seed(131)
train.index <- sample(c(1:dim(bank.df)[1]), dim(bank.df)[1]*0.6)
train.df <- bank.df[train.index, ]
valid.df <- bank.df[-train.index, ]
```

Build a classification tree
```{r}
default.ct <- rpart(Personal.Loan ~ ., data = train.df, method = "class")
```

Plot tree
```{r}
prp(default.ct, type = 1, extra = 1, under = TRUE, split.font = 1, varlen = -10)
```

Let us investigate a larger tree
```{r}
deeper.ct <- rpart(Personal.Loan ~ ., data = train.df, method = "class", cp = 0, minsplit = 1)
```

Count number of leaves
```{r}
length(deeper.ct$frame$var[deeper.ct$frame$var == "<leaf>"])
```

Plot tree
```{r}
prp(deeper.ct, type = 1, extra = 1, under = TRUE, split.font = 1, varlen = -10,
      box.col=ifelse(deeper.ct$frame$var == "<leaf>", 'gray', 'white'))
```


Compare predictions from both trees on both the training and validation data.
We need to set argument `type = "class"` in `predict()` to generate predicted class membership.


Generate confusion matrix for training data
```{r}
default.ct.point.pred.train <- predict(default.ct, train.df, type = "class")
deeper.ct.point.pred.train <- predict(deeper.ct, train.df, type = "class")
cm.default.train <- confusionMatrix(default.ct.point.pred.train, train.df$Personal.Loan)
cm.deeper.train <- confusionMatrix(deeper.ct.point.pred.train, train.df$Personal.Loan)
print(cm.default.train)
print(cm.deeper.train)
```



Generate confusion matrix for validation data
```{r}
default.ct.point.pred.valid <- predict(default.ct, valid.df, type = "class")
deeper.ct.point.pred.valid <- predict(deeper.ct, valid.df, type = "class")
cm.default.valid <- confusionMatrix(default.ct.point.pred.valid, valid.df$Personal.Loan)
cm.deeper.valid <- confusionMatrix(deeper.ct.point.pred.valid, valid.df$Personal.Loan)
print(cm.default.valid)
print(cm.deeper.valid)
```


Argument `xval` refers to the number of folds to use in rpart's built-in cross-validation procedure.
Argument `cp` sets the smallest value for the complexity parameter.

```{r}
cv.ct <- rpart(Personal.Loan ~ ., data = train.df, method = "class",
            cp = 0.00001, minsplit = 5, xval = 5)
printcp(cv.ct)
plotcp(cv.ct)
```


Prune by lower cp
```{r}
pruned.ct <- prune(cv.ct, cp = cv.ct$cptable[which.min(cv.ct$cptable[,"xerror"]),"CP"])
length(pruned.ct$frame$var[pruned.ct$frame$var == "<leaf>"])
prp(pruned.ct, type = 1, extra = 1, split.font = 1, varlen = -10)
```

Improved version of pruning 
```{r}
# this is the cp parameter with smallest cv-errror
index_cp_min = which.min(cv.ct$cptable[,"xerror"])

# one standard deviation rule 
# need to find first cp value for which the xerror is below horizontal line on the plot
(val_h = cv.ct$cptable[index_cp_min, "xerror"] + cv.ct$cptable[index_cp_min, "xstd"])
(index_cp_std = Position(function(x) x < val_h, cv.ct$cptable[, "xerror"]))
(cp_std = cv.ct$cptable[ index_cp_std, "CP" ])  
```


```{r}
pruned.ct <- prune(cv.ct, cp = cp_std)
length(pruned.ct$frame$var[pruned.ct$frame$var == "<leaf>"])
prp(pruned.ct, type = 1, extra = 1, split.font = 1, varlen = -10)
```

