---
title: "Tabloid data set"
author: ""
date: ''
output: 
    pdf_document:
        number_sections: true
        includes:
            in_header: mystyles.sty
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
options(digits=3)
```

# Description

A large retailer wants to explore the predictability of response to a
tabloid mailing.

\sk 

If they mail a tabloid to a customer in their data-base, can they
predict whether or not the customer will respond by making a
purchase.

\sk 

The dependent variable is 1 if they buy something, 0 if they do not.

\sk

They tried to come up with x’s based on past purchasing behavior.

\sk

The Predictive Analytics team builds a model for \bl the probability the customer responds \bk given \rd information about the customer\bk.

\sk

What information about a customer do they use?

- `nTab`: number of past orders.
- `moCbook`: months since last order.
- `iRecMer1`: 1/months since last order in merchandise category 1.
- `llDol`: log of the dollar value of past purchases.

The data for these variables is obtained from the companies
operational data base.


# Preprocessing

We download the data and preprocess it first

```{r}
if (!file.exists("Tabloid_test.csv"))
   download.file(
       'https://github.com/ChicagoBoothML/MLClassData/raw/master/Tabloid/Tabloid_test.csv',
       'Tabloid_test.csv')

if (!file.exists("Tabloid_train.csv"))
   download.file(
       'https://github.com/ChicagoBoothML/MLClassData/raw/master/Tabloid/Tabloid_train.csv',
       'Tabloid_train.csv')


td = read.csv("Tabloid_train.csv")
td_validation = read.csv("Tabloid_test.csv")

td$purchase = as.factor(td$purchase)
td_validation$purchase = as.factor(td_validation$purchase)
```

\newpage



# Summary statistics

```{r}
summary(td)
```

Notice that the percentage of households that make a purchase is pretty small!

\sk

$258/10000 = 0.0258$

\sk

Illustration of how `nTab` is related to `responders`.  
```{r fig.width=10, fig.height=5}
par(mfrow=c(1,2))
hist(td[td$purchase==0, "nTab"], breaks=40, col="red", 
     main="nonresponders", xlab="nTab", xlim=c(0,85))
hist(td[td$purchase==1, "nTab"], breaks=40, col="blue", 
     main="responders", xlab="nTab", xlim=c(0,85))
```

\newpage

Here is `Y` plotted vs. each of the four `X`'s

\sk
```{r fig.width=8, fig.height=8}
par(mfrow=c(2,2), mar=c(3,3,3,1), mgp=c(2,1,0)) 
plot(nTab~purchase,td,col=c("red", "blue"))
plot(moCbook~purchase,td,col=c("red", "blue"))
plot(iRecMer1~purchase,td,col=c("red", "blue"))
plot(llDol~purchase,td,col=c("red", "blue"))
```






\newpage
# Fit models

We fit

* logistic regression
* random forest model
* boosting

```{r message=FALSE}
library(caret)
library(tree)
library(ranger)
library(xgboost)
```

I am creating a list to store results of  all models.
```{r}
phat.list = list() #store the test phat for the different methods here
```


## Logistic regression

We fit a logistic regression model using all variables
```{r}
lgfit = glm(purchase~., td, family=binomial)
print(summary(lgfit))
```

Predictions are stored for later analysis.
Note that we use `type="response"` which specifies the  type of prediction used. 
For a default binomial model the default predictions are of log-odds (probabilities on logit scale).
By using `type = "response"`, we obtain the predicted probabilities.
```{r}
phat = predict(lgfit, td_validation, type="response")
phat.list$logit = matrix(phat,ncol=1) 
```

\newpage

## Random Forest

We fit random forest models for a few different settings.

```{r}
p=ncol(td)-1

hyper_grid_rf <- expand.grid(
  mtry       = c(p, ceiling(sqrt(p))),
  node_size  = c(5, 10, 20)
)

# we will store phat values here
phat.list$rf = matrix(0.0, nrow(td_validation), nrow(hyper_grid_rf))  

for(i in 1:nrow(hyper_grid_rf)) {
  # train model
  rf.model <- ranger(
    formula         = purchase~.,
    data            = td, 
    num.trees       = 250,
    mtry            = hyper_grid_rf$mtry[i],
    min.node.size   = hyper_grid_rf$node_size[i],
    probability     = TRUE, 
    seed            = 99
  )   
   
   # predict for random forest returns
   # a matrix of class probabilities 
   #    one column for each class and one row for each input
   # we want to record probability for class=1, 
   # which is the second column of the output
   phat = predict(rf.model, data=td_validation)$predictions[,2]
   phat.list$rf[,i]=phat
}
```

\newpage

## Boosting


Remember that we need to put our data into a suitable form.

```{r}
X.train = as.matrix( td[,-1] )
Y.train = as.numeric(td$purchase)-1
X.validation = as.matrix( td_validation[,-1] )
Y.validation = as.numeric(td_validation$purchase) - 1
```

We fit boosting models for a few different settings.

```{r}
hyper_grid_xgb <- expand.grid(
  shrinkage = c(.01, .1),        ## controls the learning rate
  interaction.depth = c(1, 2, 4), ## tree depth
  nrounds = c(1000, 3000)         ## number of trees
)

# we will store phat values here
phat.list$boost = matrix(0.0,nrow(td_validation),nrow(hyper_grid_xgb))
```


Fitting
```{r message=F, warning=F, error=F}
for(i in 1:nrow(hyper_grid_xgb)) {
  # create parameter list
  params <- list(
    eta = hyper_grid_xgb$shrinkage[i],
    max_depth = hyper_grid_xgb$interaction.depth[i]
  )
   
  # reproducibility
  set.seed(4776)
  
  # train model
  xgb.model <- xgboost(
    data      = X.train,
    label     = Y.train,
    params    = params,
    nrounds   = hyper_grid_xgb$nrounds[i],
    objective = "binary:logistic",     # for regression models
    verbose   = 0,                     # silent
    verbosity = 0                      # silent
  )
   
  phat = predict(xgb.model, newdata=X.validation)
  phat.list$boost[,i] = phat
}
```

\newpage

# Analysis of results

## Misclassification rate

Let us first look at misclassification rate.

The following function computes the confusion matrix
from the vector of true class labels `y` and a vector
of estimated probabilities. The probabilities are converted
into predicted class labels using a threshold `thr`.

```{r}
# y should be 0/1
# phat are probabilities obtained by our algorithm 
# thr is the cut off value - everything above thr is classified as 1
getConfusionMatrix = function(y,phat,thr=0.5) {
   yhat = as.factor( ifelse(phat > thr, 1, 0) )
   confusionMatrix(yhat, y)
}
```

This function computes the misclassification rate
from the vector of true class labels `y` and a vector
of estimated probabilities. The probabilities are converted
into predicted class labels using a threshold `thr`.

```{r}
# y should be 0/1
# phat are probabilities obtained by our algorithm 
# thr is the cut off value - everything above thr is classified as 1
loss.misclassification.rate = function(y, phat, thr=0.5) 
   1 - getConfusionMatrix(y, phat, thr)$overall[1]
```

For **logistic regression** we have:
```{r}
cfm <- getConfusionMatrix(td_validation$purchase, phat.list$logit[,1], 0.5)
print(cfm, printStats = F)
cat('misclassification rate = ', 
    loss.misclassification.rate(td_validation$purchase, phat.list[[1]][,1], 0.5), 
    '\n')
```

\newpage

For **random forest** we have:
```{r}
nrun = nrow(hyper_grid_rf)
for(j in 1:nrun) {
  print(hyper_grid_rf[j,])
  cfm <- getConfusionMatrix(td_validation$purchase, phat.list[[2]][,j], 0.5)   
  print(cfm, printStats = F)
  cat('misclassification rate = ', 
      loss.misclassification.rate(td_validation$purchase, phat.list[[2]][,j], 0.5), 
      '\n')
}
```

\newpage

For **boosting** we have:
```{r}
nrun = nrow(hyper_grid_xgb)
for(j in 1:nrun) {
  print(hyper_grid_xgb[j,])
  cfm <- getConfusionMatrix(td_validation$purchase, phat.list[[3]][,j], 0.5)
  print(cfm, printStats = F)
  cat('misclassification rate = ', 
      loss.misclassification.rate(td_validation$purchase, phat.list[[3]][,j], 0.5), 
      '\n')
}
```

\newpage


This is strange... There seems to be fit in the model.
```{r fig.width=4, fig.height=4}
par(mar=c(3,3,3,1), mgp=c(2,1,0)) 
phat = predict(lgfit, newdata=td, type="response")
plot(phat~td$purchase, col=c("red","blue"), 
     xlab="purchase", ylab="phat", ylim=c(0,1.05), cex.text=0.7)
```


\newpage

The idea behind the tabloid example is that if we can predict who
will buy we can target those customers and send them the tabloid.


To get an idea of how well our model is working, we can imagine
choosing a customer from the data set to mail to first - did they
buy?

We can look at the y value to see if they bought.


Whom would you mail to first?

You could mail the first 40 people in your database.

\small
```{r}
td$phat = phat
td[1:40, c("purchase", "phat")]
```

\normalsize \sk

Out of the first 40, there is only one purchase.

\newpage

If you believe your model, you might mail to the household with
the largest $\hat p$ (estimated prob of buying) first. 
Then you would mail to the household with the second largest $\hat p$
and so on.


\small
```{r}
td$phat = phat
sorted_phat = order(-phat)
td[sorted_phat[1:40], c("purchase", "phat")]
```

\normalsize
\sk
You got 16 purchases out of the first 40 customers you targeted.
Using only 40/10000 = 0.004 of the data we got 16/258 = .062 of
the purchases!









\newpage

## Deviance

The following function is used to compute the deviance of a model.

```{r}
# deviance loss function
# y should be 0/1
# phat are probabilities obtained by our algorithm 
# wht shrinks probabilities in phat towards .5 
#   this helps avoid numerical problems --- don't use log(0)!
lossf = function(y,phat,wht=0.0000001) {
   if(is.factor(y)) y = as.numeric(y)-1
   phat = (1-wht)*phat + wht*.5
   py = ifelse(y==1, phat, 1-phat)
   return(-2*sum(log(py)))
}
```


Plot validation set loss --- deviance:

```{r fig.width=8, fig.height=8}
lossL = list()
nmethod = length(phat.list)
for(i in 1:nmethod) {
   nrun = ncol(phat.list[[i]])
   lvec = rep(0,nrun)
   for(j in 1:nrun) lvec[j] = lossf(td_validation$purchase, phat.list[[i]][,j])
   lossL[[i]]=lvec; names(lossL)[i] = names(phat.list)[i]
}
lossv = unlist(lossL)
plot(lossv, ylab="out-of-sample loss", type="n")
nloss=0
for(i in 1:nmethod) {
   ii = nloss + 1:ncol(phat.list[[i]])
   points(ii,lossv[ii],col=i,pch=17)
   nloss = nloss + ncol(phat.list[[i]])
}
legend("topright",legend=names(phat.list),col=1:nmethod,pch=rep(17,nmethod))
```

From each method class, we choose the one that has the lowest error on the validation set.

```{r}
nmethod = length(phat.list)
phatBest = matrix(0.0,nrow(td_validation),nmethod) #pick off best from each method
colnames(phatBest) = names(phat.list)
for(i in 1:nmethod) {
   nrun = ncol(phat.list[[i]])
   lvec = rep(0,nrun)
   for(j in 1:nrun) lvec[j] = lossf(td_validation$purchase,phat.list[[i]][,j])
   imin = which.min(lvec)
   phatBest[,i] = phat.list[[i]][,imin]
}
```



\newpage

## Expected value of a classifier

Let us target everyone with $\hat p > 0.02$

Our **cost/benefit matrix** looks like this
```{r}
cost_benefit = matrix(c(0,-0.8,0,39.20), nrow=2)
print(cost_benefit)
```

Expected values of targeting is below:

```{r}
confMat = getConfusionMatrix(td_validation$purchase, phatBest[,1], 0.02)
print(confMat, printStats = F)
cat("Expected value of targeting using logistic regression = ", 
    sum(sum(as.matrix(confMat) * cost_benefit)), "\n")
```

```{r}
confMat = getConfusionMatrix(td_validation$purchase, phatBest[,2], 0.02)
print(confMat, printStats = F)
cat("Expected value of targeting using random forests = ", 
    sum(sum(as.matrix(confMat) * cost_benefit)), "\n")
```

```{r}
confMat = getConfusionMatrix(td_validation$purchase, phatBest[,3], 0.02)
print(confMat, printStats = F)
cat("Expected value of targeting using boosting = ", 
    sum(sum(as.matrix(confMat) * cost_benefit)), "\n")
```

\newpage

## ROC curves

Library for plotting various summary curves
```{r}
library(ROCR)
```

```{r fig.width=6, fig.height=6}
for(i in 1:ncol(phatBest)) {
   pred = prediction(phatBest[,i], td_validation$purchase)
   perf = performance(pred, measure = "tpr", x.measure = "fpr")
   
   if (i == 1) {
     plot(perf, col = 1, lwd = 2,
          main= 'ROC curve', xlab='FPR', ylab='TPR', cex.lab=1)
   } else {
     plot(perf, add = T, col = i, lwd = 2)
   }
}
abline(0,1,lty=2)
legend("topleft",legend=names(phat.list),col=1:nmethod,lty=rep(1,nmethod))
```

We can also compute AUC (area under the ROC curve).

```{r}
for(i in 1:ncol(phatBest)) {
  pred = prediction(phatBest[,i], td_validation$purchase)
  perf <- performance(pred, measure = "auc")
  print(paste0("AUC ", names(phat.list)[i], " :: ", perf@y.values[[1]]))
}
```


\newpage 

## Lift curves

```{r fig.width=6, fig.height=6}
pred = prediction(phatBest[,1], td_validation$purchase)
perf = performance(pred, measure = "lift", x.measure = "rpp", lwd=2)
plot(perf, col=1, ylim=c(0,5))
abline(h=1, lty=2)

for(i in 2:ncol(phatBest)) {
   pred = prediction(phatBest[,i], td_validation$purchase)
   perf = performance(pred, measure = "lift", x.measure = "rpp")
   plot(perf, add = T, col = i, lwd = 2)
}
legend("topright",legend=names(phat.list),col=1:nmethod,lty=rep(1,nmethod), lwd=2)
```

\newpage

## Cummulative response 


```{r fig.width=6, fig.height=6}
pred = prediction(phatBest[,1], td_validation$purchase)
perf = performance(pred, measure = "tpr", x.measure = "rpp")
plot(perf, col=1, ylim=c(0,1),lwd=2)
abline(h=1, lty=2)
abline(0,1,lty=2)
for(i in 2:ncol(phatBest)) {
   pred = prediction(phatBest[,i], td_validation$purchase)
   perf = performance(pred, measure = "tpr", x.measure = "rpp")
   plot(perf, add = T, col = i, lwd = 2)
}
legend("bottomright",legend=names(phat.list),col=1:nmethod,lty=rep(1,nmethod),lwd=2)
```


