---
title: "Classification Example"
output: 
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
options(digits=3)
options(width = 48)

library(dplyr)
library(ranger)
library(iml)
library(ggplot2)

set.seed(1)
```

We are going to use the data from:

- _Fernandes, Cardoso, and Fernandes_. "Transfer learning with partial observability applied to cervical cancer screening." In Iberian Conference on Pattern Recognition and Image Analysis, 243–50. Springer. (2017). [URL](https://link.springer.com/chapter/10.1007/978-3-319-58838-4_27)

The original data is available from 
the [UCI Machine learning repository](https://archive.ics.uci.edu/ml/datasets/Cervical+cancer+%28Risk+Factors%29).
We are using the preprocessed data,
which was obtained using the script [here](https://github.com/christophM/interpretable-ml-book/blob/master/R/get-cervical-cancer-dataset.R).


We are going to predict the `Biopsy` variable,
which serves as the gold standard for diagnosing cervical cancer. 


Load the data
```{r}
cervical = read.csv('cervical.csv')
cervical$Biopsy = as.factor(cervical$Biopsy)
```

```{r}
rf.fit <- ranger(
  formula = Biopsy ~ .,
  data            = cervical, 
  num.trees       = 250,
  importance      = 'impurity',
  probability = T
)
```


```{r warning=F, message=F, error=F}
# Create custom predict function that returns the predicted values as a vector 
pred.rf.fun <- function(model, newdata)  
  predict(model, data = newdata)$predictions

mod <- Predictor$new(rf.fit, 
                     data = cervical,
                     class = "Cancer",
                     predict.fun = pred.rf.fun    # we need a custom predict function
                     )
pdp = FeatureEffect$new(mod, "Age", method = "pdp")
p1 = pdp$plot(ylim=c(0, 0.4)) + 
  scale_x_continuous(limits = c(0, NA)) +
  scale_y_continuous('Predicted cancer probability', limits=c(0, 0.4))

pdp$set.feature("Hormonal.Contraceptives..years.")
p2 = pdp$plot(ylim=c(0, 0.4)) + 
  scale_x_continuous("Years on hormonal contraceptives", limits = c(0, NA)) +
  scale_y_continuous('', limits=c(0, 0.4))
gridExtra::grid.arrange(p1, p2, ncol = 2)
```


```{r warning=F, message=F, error=F}
cervical_subset_index = sample(1:nrow(cervical), size = 300)
cervical_subset = cervical[cervical_subset_index, ]

pred.cervical = Predictor$new(rf.fit, 
                              data = cervical_subset, 
                              class = "Cancer",
                              predict.fun = pred.rf.fun
                              )
ice = FeatureEffect$new(pred.cervical, 
                        "Age", 
                        center.at = min(cervical_subset$Age),
                        method = "pdp+ice")$plot() + 
  scale_color_discrete(guide='none') + 
  scale_y_continuous('Predicted cancer probability')
ice
```


PDP of cancer probability and the interaction of age and number of pregnancies.

```{r}
pd = FeatureEffect$new(mod, c("Age", "Num.of.pregnancies"), method = "pdp") 
pd$plot() 
```




