---
title: "Hockey analytics"
author: "Mladen Kolar (mkolar@chicagobooth.edu)"
date: ''
fontsize: 10pt
output:
  html_document: default    
  pdf_document: 
      includes:
        in_header: mystyles.sty  
subtitle: "Finding good players using variable selection"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
options(digits=3)
options(width = 100)
options(warn = -1)
```


We are going to investigate data on all of the goals in the
2002–2014 seasons of the National Hockey League (NHL).

- See _Robert Gramacy, Matt Taddy, and Sen Tian_. **Hockey performance via regression**. _Handbook
of Statistical Methods for Design and Analysis in Sports_, 2015. For more details.


The data is available in the `gamlr` package, a competitor of `glmnet`.

```{r message=FALSE}
library(gamlr)
data(hockey)
```

- The data was scraped from NHL.com using an R package called `nhlscrapr`, spanning 11 seasons: 2002-03 through 2013--14, with playoffs.
- It includes other info we're not going to use (shots, blocked shots, penalties, etc.)



Hockey is like soccer, but on ice, 6-on-6 and with **rapid substitution**.

\sk

\includegraphics[width=0.9\linewidth]{images/hockey.png}

\sk

Quantifying player performance in hockey is hard:

- continuous nature of play
- infrequent number of goals
- combinatorially huge numbers of player configurations.

One popular metric of individual player performance is **plus--minus**:

- the number of goals scored by the player's team,
- minus the number scored by the opposing team

while that player is on the ice.


Plus-minus is better than just goals, because it distributes the credit and blame.

\sk

\includegraphics[width=0.9\linewidth]{images/hockey_pm.png}

\sk

The limits of this approach are obvious: there is no accounting for teammates or opponents. 

In hockey, where players tend to be grouped together on "lines" and coaches will 
"line match" against opponents, a player's PM can be artificially 
inflated or deflated by the play of his opponents and peers.

\sk

In summary, two disadvantages to plus--minus:

- It is a **marginal effect**, averaging over situation, say.
- It doesn't control for sample size.

\sk

A better measure of performance would be a **partial effect**, having controlled for the effect of 

- teammates,
- opponents, etc.

An appealing aspect of such an analysis is that it requires no extra data beyond that used to calculate plus--minus,

- just a (much) more involved calculation.

\sk

We will build a better performance metric with regression.



# The setup


```{r}
head(goal)
```


Given $n$ goals throughout the National Hockey League (NHL) over some specified time
period, say 

$$
y_i = \left\{\begin{array}{cl}
+1 & \mbox{ for a goal by the home team,  and} \\
-1 & \mbox{ for a goal by the away team.}
\end{array}\right.
$$

Then, say that 

$$
q_i = \mathbb{P}(Y_i = 1) = \mathbb{P}(\mbox{home team scored goal $i$}).
$$

- *Home* and *away* are merely organizational devices, creating a consistent binary bifurcation for goals that can be applied across games, seasons, etc. 

- Due to the symmetry in the logit transformation, player effects are unchanged when framing away team probabilities as $q_i$ rather than $(1 - q_i)$, so we loose no generality by "privileging" home team goals in this way.


**Player model**

The simplest version of a model for partial player effects is the so-called **player model**, where

- the **log odds** that the home team has scored a given goal, $i$, becomes

$$
\log\left( \frac{q_i}{1 - q_i} \right) = \alpha + \beta_{h_{i_1}} + \cdots + \beta_{h_{i_6}} - \beta_{a_{i_1}} - \cdots - \beta_{a_{i_6}},
$$

where 

- $h_{i_1}, \dots, h_{i_6}$ are the home team's players (i.e., player indicators), and
- $a_{i_1}, \dots, a_{i_6}$ are the away team players.  


The coefficients $\beta_*$ are our **partial player effects**!

- What does $\alpha$ represent?


**The data**

How do we set up the data so that it is faithful to this format, in a logistic regression setup?  Like this:

\sk

\includegraphics[width=0.9\linewidth]{images/playermodel.png}


- Notice that the design matrix $X_P$ is **sparse**.
- Sparse matrix libraries can ease storage and computational burden.

```{r}
player[1:3, 2:7]
```

\sk

**Getting fancier**

Beyond controlling for the effect of who else is on the ice, we also want
to control for things unrelated to player ability.

Embellishments abound.  You can add:

- player-season indicators;
- Team or team--season indicators;
- Special teams indicators (6v5, 6v4, ..., pulled goalie, etc.);
- Special situations: overtime, playoffs, exhibition, etc.

But the idea is the same:

- These are just indicator variables,
- and its all just a big logistic regression.


# `gamlr` package

We will use `gamlr` package to fit the model

- `gamlr` preferrs a so-called log-gamma penaty ${\rm pen}(\beta) = \log(1 + |\beta)|)$.

\sk

\includegraphics[width=0.9\linewidth]{images/penalties.png}

\sk

It works very similarly to `glmnet`, but are some small differences. E.g., it

- supports selection via information criteria, in addition to CV;
- allows some coefficients to undergo different (e.g., ridge/no) penalization.

\sk\sk

**Differential penalization**

\sk

This ability to differentially penalize could be advantageous.

\sk

If (large) player partial effects (assets and liabilities) are the main interest,

- i.e., large non-zero coefficients, separating the wheat from the chaff,

then it makes sense to "select" players, but be more lenient to special teams, etc.

\sk

Ok, we're sold.


# Assembling the design

Lets throw everything together:

- `config` with special teams, playoff indicators, etc.
- `team` with team indicators, and 
- `player` with player indicators

These are sparse matrices, so we'll need to combine them together using a new command.

```{r}
X <- cbind(config, team, player)
y <- goal$homegoal
dim(X)
```

- Woah!  That's quite big.


## `gamlr` call

```{r}
nhlreg <- gamlr(X, y, free=1:(ncol(config)+ncol(team)), 
                 family="binomial", standardize=FALSE)
```

- `free` denotes unpenalized columns. These are columns of the design matrix 
that we do not want penalized. We are using it to keep the special-teams
and team-season variables unpenalized---we know that we want them in the model,
and so we let them enter without restriction.

- We use `standardize=FALSE` because the columns are already indicators.
  This is one of the special cases where all of our penalized variables
  are on the same scale (player presence or absence). Without `standardize=FALSE`,
  we would be multiplying the penalty for each coefficient (player effect) 
  by that player's standard deviation in the player matrix. 
  The players with big standard deviation are guys who play a lot. 
  Players with small standard deviation are those who play little 
  (almost all zeros). Hence, weighting penalty by standard deviations 
  in this case is exactly what we do not want: a bigger penalty for people
  with many minutes on ice, a smaller penalty for those who seldom play. 
  Indeed, running the regression without `standardize=FALSE`
  leads to a bunch of farm-team players coming up on top.


Now how are we going to look at this output, with nearly 3000 coefficients? Patiently.

\sk

Start with $\hat{\alpha}$, the home-ice advantage (ignoring everything else).

```{r}
exp(coef(nhlreg)[1])
```

- Home ice increases the odds that the home team has scored by 8%.
  Without conditioning on any of the other covariates, the home team is around 8% more
  likely to have scored any given goal. That is a big home-ice advantage!


## (De-) Selected players

Lets extract the coefficients.

```{r}
Baicc <- coef(nhlreg)[colnames(player),]
```

By default, the reported coefficients are from the best model by **AICc**,

- a ["corrected" AIC](https://en.wikipedia.org/wiki/Akaike_information_criterion#AICc) criterion.

\sk

How many are non-zero

```{r}
c(nonzero=sum(Baicc != 0), prop=mean(Baicc != 0), 
  assets=mean(Baicc > 0), liabilities=mean(Baicc < 0))
```

- About 75% of the league is "average".
- 16% assets, 10% liabilities.


## Top/bottom ten

Here are the top ten players. They are almost all recognizable stars.

```{r}
Baicc[order(Baicc, decreasing=TRUE)[1:10]]
```


And the bottom ten. They are not those with little ice
time, but rather those with much ice time who underperform.


```{r}
Baicc[order(Baicc)[1:10]]
```


Let us compare to what would happen if we run the 
regression without `standardize=FALSE`.

```{r}
nhlreg.std <- gamlr(X, y, free=1:(ncol(config)+ncol(team)), 
                 family="binomial")
Baicc.std <- coef(nhlreg.std)[colnames(player),]
Baicc.std[order(Baicc.std, decreasing=TRUE)[1:10]]
```



## Contribution to goal for/against

Whenever a goal is scored, 

- Pittsburg's odds of having scored (rather than being scored on) increase by 51% if Sidney Crosby is on the ice;

```{r}
exp(Baicc["SIDNEY_CROSBY"])
```

- and the Blue Jackets' (or Kings', pre 2011-12) odds of having scored drop by 22% if Jack Johnson is on the ice.

```{r}
exp(Baicc["JACK_JOHNSON"])
```

(Remember, the data is a little old.)





# Cross-validation

Cross-validation results instead?

```{r cache=TRUE, fig.align="center"}
cv.nhlreg <- cv.gamlr(X, y, free=1:(ncol(config)+ncol(team)), family="binomial", 
  standardize=FALSE)
cv.nhlreg
```

The `cv.gamlr` object stores a `gamlr` object (the full data path fit) as one
of its entries, and you can plot both the regularization paths and the CV
experiment.

```{r}
par(mfrow=c(1,2)); plot(cv.nhlreg); plot(cv.nhlreg$gamlr) 
```


Let us look at $\log(\hat{\lambda})$ under various criteria.


```{r}
c(AICc=as.numeric(log(nhlreg$lambda[which.min(AICc(nhlreg))])), 
  AIC=as.numeric(log(nhlreg$lambda[which.min(AIC(nhlreg))])),
  BIC=as.numeric(log(nhlreg$lambda[which.min(BIC(nhlreg))])), 
  CVmin=log(cv.nhlreg$lambda.min), CV1se=log(cv.nhlreg$lambda.1se))
```

Lets compare de-selection to what we got with AICc.


```{r}
Bcvmin <- coef(cv.nhlreg, select="min")[colnames(player),]
Bcv1se <- coef(cv.nhlreg)[colnames(player),]
Bbic <- coef(nhlreg,select=which.min(BIC(nhlreg)))[colnames(player),]
c(AICc=sum(Baicc!=0), CVmin=sum(Bcvmin!=0), CV1se=sum(Bcv1se!=0), BIC=sum(Bbic!=0))
``` 

- Woah!  BIC way over-penalizes.



#  Partial plus-minus

Consider the situation where you have no information beyond the fact that player "$k$" is on the ice.

- All other coefficients are effectively zero.

In isolation, player $k$'s effect is the number of goals he was on the ice for, $N_k$, times 

$$
P_k - (1 - P_k) = \mathrm{P}(\mbox{scored}) - \mathrm{P}(\mbox{scored on}).
$$

- I.e., his expected "goals for" in isolation is $P_k N_k$,
- and his expected "goals against" in isolation is $N_k (1 - P_k)$.

So a **partial plus-minus (PPM)** could be defined as

$$
\mathrm{PPM}_k = N_k P_k - N_k(P_k - 1) = N_k(2 P_k -1).
$$

- which will be on the same scale as plus-minus (PM),
- and that could help if you're not good at thinking about "log odds".


## PPM calculation

Calculating PPM, and showing first 20.

```{r}
P <- exp(Baicc)/(1+exp(Baicc))
N <- colSums(abs(player))
PPM <- N*(2*P-1)
sort(PPM, decreasing=TRUE)[1:20]
```


## PM comparison

Calculating PM for comparison, and showing first 20.

- $+1$ for a goal by your team, $-1$ for a goal against.


```{r}
PM <- colSums(player*c(-1,1)[y+1])
names(PM) <- colnames(player)
sort(PM, decreasing=TRUE)[1:20] # all goalies
```


```{r}
bigs <- which(abs(PM)>200|abs(PPM)>200)
plot(PM[bigs],PPM[bigs],type="n", xlim=range(PM)*1.05, xlab="PM", ylab="PPM")
text(PM[bigs],PPM[bigs],labels=colnames(player)[bigs], cex=0.75); abline(a=0,b=1)
```



# If you're interested ...

If you want to read more, check out

- Original paper in the Journal of Quantitative Analysis in Sports.
    + [arXiv version](https://arxiv.org/abs/1209.5026)
    + Describes optimal line formation, and cost-benefit analysis via salary.
- Book chapter in the [Handbook of Statistical Methods and Analyses in Sports](https://www.amazon.com/Handbook-Statistical-Methods-Analyses-Handbooks/dp/1498737366).
    + [arXiv version](https://arxiv.org/abs/1510.02172)

\sk

Not sure if you'll see PPM on ESPN and time soon.














